---
title: "D3"
output:
  html_document:
    df_print: paged
date: "2025-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r data-load}
#setwd("/dades/eric.diez/6Q/ADEI")
setwd("C:/Users/Ericd/Pictures/ADEI")
dd <- read.csv("adult_def.csv", stringsAsFactors = TRUE);
names(dd)
dim(dd)
summary(dd)
set.seed(123)
samp<-sample(48842,5000)
par(mar = c(5, 4, 4, 2))  # aumenta espacio inferior


dd<-dd[samp,]
#set a list of numerical variables
names(dd)
attach(dd)
```


```{r hierarchical-clustering}

#euclidean distance si totes son numeriques
dcon<-data.frame (age,edu_num,cap_gain,cap_loss,hours_week)

d  <- dist(dcon[1:6,])

#move to Gower mixed distance to deal 
#simoultaneously with numerical and qualitative data

library(cluster)

#dissimilarity matrix
#do not include in actives the identifier variables nor the potential response variable

actives <- c("age", "workclass", "edu_num",
             "marital", "occupation", "relationship", "race", "sex",
             "cap_gain", "cap_loss", "hours_week", "native_country")
actives <- c(1:2,4:13)
dissimMatrix <- daisy(dd[,actives], metric = "gower", stand=TRUE)
distMatrix <- as.dist(dissimMatrix^2)

h1 <- hclust(distMatrix,method="ward.D2")  # NOTICE THE COST
#versions noves "ward.D" i abans de plot: par(mar=rep(2,4)) si se quejara de los margenes del plot

plot(h1)

k<-3

c2 <- cutree(h1, k=k)
```


# Añadir cluster al dataset
```{r add-visualize-clusters}
dd$cluster <- as.factor(c2)

# Análisis descriptivo de los clusters
table(dd$cluster)


# LETS SEE THE PARTITION VISUALLY

c1<-c2
# Basic scatter plots for three key numeric variables (modified version)
plot(age, hours_week, col=c1, main="Clusters by age and hours per week")
legend("topright", paste("cluster", 1:k), pch=1, col=1:k, cex=0.6)

plot(edu_num, hours_week, col=c1, main="Clusters by education level and hours per week")
legend("topright", paste("cluster", 1:k), pch=1, col=1:k, cex=0.6)

plot(age, edu_num, col=c1, main="Clusters by age and education level")
legend("topright", paste("cluster", 1:k), pch=1, col=1:k, cex=0.6)


pairs(dcon[,1:5], col=c1)
```


# LETS SEE THE QUALITY OF THE HIERARCHICAL PARTITION
```{r quality-profiling}
cdg <- aggregate(as.data.frame(dcon), list(c1), mean)
cdg

plot(cdg[,1], cdg[,6])
```


```{r test-functions}
#Profiling plots

#Calcula els valor test de la variable Xnum per totes les modalitats del factor P
ValorTestXnum <- function(Xnum,P){
  #freq dis of fac
  nk <- as.vector(table(P)); 
  n <- sum(nk); 
  #mitjanes x grups
  xk <- tapply(Xnum,P,mean);
  #valors test
  txk <- (xk-mean(Xnum))/(sd(Xnum)*sqrt((n-nk)/(n*nk))); 
  #p-values
  pxk <- pt(txk,n-1,lower.tail=F);
  for(c in 1:length(levels(as.factor(P)))){if (pxk[c]>0.5){pxk[c]<-1-pxk[c]}}
  return (pxk)
}




ValorTestXquali <- function(P,Xquali){
  taula <- table(P,Xquali);
  n <- sum(taula); 
  pk <- apply(taula,1,sum)/n;
  pj <- apply(taula,2,sum)/n;
  pf <- taula/(n*pk);
  pjm <- matrix(data=pj,nrow=dim(pf)[1],ncol=dim(pf)[2], byrow=TRUE);      
  dpf <- pf - pjm; 
  dvt <- sqrt(((1-pk)/(n*pk))%*%t(pj*(1-pj))); 
  #i hi ha divisions iguals a 0 dona NA i no funciona
  zkj <- dpf
  zkj[dpf!=0]<-dpf[dpf!=0]/dvt[dpf!=0]; 
  pzkj <- pnorm(zkj,lower.tail=F);
  for(c in 1:length(levels(as.factor(P)))){for (s in 1:length(levels(Xquali))){if (pzkj[c,s]> 0.5){pzkj[c,s]<-1- pzkj[c,s]}}}
  return (list(rowpf=pf,vtest=zkj,pval=pzkj))
}
```

```{r descriptive-analysis}
#source("file")
#dades contain the dataset
dades <- dd[, setdiff(names(dd),
                      c("fnlwgt", "income", "cluster", "income_integer"))]
#dades<-dd[filtro,]
#dades<-df
K<-dim(dades)[2]
par(ask=TRUE)


#P must contain the class variable
#P<-dd[,3]
P<-c2
#P<-dd[,18]
nameP<-"classe"
#P<-df[,33]

nc<-length(levels(factor(P)))
nc
pvalk <- matrix(data=0,nrow=nc,ncol=K, dimnames=list(levels(P),names(dades)))
nameP<-"Class"
n<-dim(dades)[1]

for(k in 1:K){
  if (is.numeric(dades[,k])){ 
    print(paste("Anàlisi per classes de la Variable:", names(dades)[k]))
    
    boxplot(dades[,k]~P, main=paste("Boxplot of", names(dades)[k], "vs", nameP ), horizontal=TRUE)
    
    barplot(tapply(dades[[k]], P, mean),main=paste("Means of", names(dades)[k], "by", nameP ))
    abline(h=mean(dades[[k]]))
    legend(0,mean(dades[[k]]),"global mean",bty="n")
    print("Estadístics per groups:")
    for(s in levels(as.factor(P))) {print(summary(dades[P==s,k]))}
    o<-oneway.test(dades[,k]~P)
    print(paste("p-valueANOVA:", o$p.value))
    kw<-kruskal.test(dades[,k]~P)
    print(paste("p-value Kruskal-Wallis:", kw$p.value))
    pvalk[,k]<-ValorTestXnum(dades[,k], P)
    print("p-values ValorsTest: ")
    print(pvalk[,k])      
  }else{
    if(class(dd[,k])=="Date"){
      print(summary(dd[,k]))
      print(sd(dd[,k]))
      #decide breaks: weeks, months, quarters...
      hist(dd[,k],breaks="weeks")
    }else{
      #qualitatives
      print(paste("Variable", names(dades)[k]))
      table<-table(P,dades[,k])
      #   print("Cross-table")
      #   print(table)
      rowperc<-prop.table(table,1)
      
      colperc<-prop.table(table,2)
      #  print("Distribucions condicionades a files")
      # print(rowperc)
      
      #ojo porque si la variable es true o false la identifica amb el tipus Logical i
      #aquest no te levels, por tanto, coertion preventiva
      
      dades[,k]<-as.factor(dades[,k])
      
      
      marg <- table(as.factor(P))/n
      print(append("Categories=",levels(as.factor(dades[,k]))))
      par(mar = c(8, 4, 4, 2))  # más espacio abajo
      
    
      #from next plots, select one of them according to your practical case
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(colperc[,c],col=paleta[c]) }
      
      #with legend
      par(mar = c(5, 4, 4, 6)) 
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(colperc[,c],col=paleta[c]) }
      legend("topright",inset = c(-0.19, 0), levels(dades[,k]), col=paleta, lty=2, cex=0.6, bg = "transparent", box.lwd = 0, xpd = TRUE)
      par(mar = c(8, 4, 4, 2)) 
      
      #condicionades a classes
      print(append("Categories=",levels(dades[,k])))
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(rowperc[,c],col=paleta[c]) }
      
      #with legend
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(rowperc[,c],col=paleta[c]) }
      legend("topright", levels(dades[,k]), col=paleta, lty=2, cex=0.6)
      
      #amb variable en eix d'abcisses
      marg <-table(dades[,k])/n
      print(append("Categories=",levels(dades[,k])))
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]), las=3)
      #x<-plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]), xaxt="n")
      #text(x=x+.25, y=-1, adj=1, levels(CountryName), xpd=TRUE, srt=25, cex=0.7)
      paleta<-rainbow(length(levels(as.factor(P))))
      for(c in 1:length(levels(as.factor(P)))){lines(rowperc[c,],col=paleta[c]) }
      
      #with legend
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]), las=3)
      for(c in 1:length(levels(as.factor(P)))){lines(rowperc[c,],col=paleta[c])}
      legend("topright", levels(as.factor(P)), col=paleta, lty=2, cex=0.6)
      
      #condicionades a columna 
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]), las=3)
      paleta<-rainbow(length(levels(as.factor(P))))
      for(c in 1:length(levels(as.factor(P)))){lines(colperc[c,],col=paleta[c]) }
      
      #with legend
      par(mar = c(8, 4, 4, 6)) 
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of classes by",names(dades)[k]), las=3)
      for(c in 1:length(levels(as.factor(P)))){lines(colperc[c,],col=paleta[c])}
      legend("topright", inset = c(-0.10, 0),levels(as.factor(P)), col=paleta, lty=2, cex=0.6, bg = "transparent", box.lwd = 0, xpd = TRUE)
      par(mar = c(8, 4, 4, 2)) 
      
      table<-table(dades[,k],P)
      print("Cross Table:")
      print(table)
      print("Distribucions condicionades a columnes:")
      print(colperc)
      
      #diagrames de barres apilades                                         
      
      paleta<-rainbow(length(levels(dades[,k])))
      barplot(table(dades[,k], as.factor(P)), beside=FALSE,col=paleta )
      
      barplot(table(dades[,k], as.factor(P)), beside=FALSE,col=paleta )
      legend("topright",levels(as.factor(dades[,k])),pch=1,cex=0.5, col=paleta)
      
      #diagrames de barres adosades
      barplot(table(dades[,k], as.factor(P)), beside=TRUE,col=paleta )
      
      par(mar = c(5, 4, 4, 6)) 
      barplot(table(dades[,k], as.factor(P)), beside=TRUE,col=paleta)
      legend("topright", inset = c(-0.17, 0),levels(as.factor(dades[,k])),pch=1,cex=0.5, col=paleta, bg = "transparent", box.lwd = 0, xpd = TRUE)
      par(mar = c(8, 4, 4, 2)) 
      
      print("Test Chi quadrat: ")
      print(chisq.test(dades[,k], as.factor(P)))
      
      print("valorsTest:")
      print( ValorTestXquali(P,dades[,k]))
      #calcular els pvalues de les quali
    }
  }
}#endfor

#descriptors de les classes més significatius. Afegir info qualits
for (c in 1:length(levels(as.factor(P)))) {
  if(!is.na(levels(as.factor(P))[c])){
    print(paste("P.values per class:",levels(as.factor(P))[c]));
    print(sort(pvalk[c,]), digits=3) 
  }
}
```

#PCA
# 1. Load Libraries and Data

```{r load_data_PCA}
library(dplyr) 
setwd("C:/Users/Ericd/Pictures/ADEI")
dd <- read.csv("adult_def.csv", stringsAsFactors = TRUE);
dd$income_per_hour <- dd$income_integer / dd$hours_week
set.seed(123)
samp <- sample(48842, 5000)
dd <- dd[samp,]
objects()
attributes(dd)
# Remove unwanted columns
dd <- dd[, !(names(dd) %in% c("fnlwgt", "edu_num"))]
```


# 2. Initial Exploration

```{r exploration}
# DATA VISUALIZATION
attach(dd)
names(dd)

# Check classes of variables
sapply(dd, class)
```

# 3. Selection of Continuous Variables

```{r continuous_variables}
# List of numeric variables
numeriques <- which(sapply(dd, is.numeric))
numeriques

# Create data frame with continuous variables
dcon <- dd[, numeriques]
sapply(dcon, class)

# Note: If there were missing data, one should handle it before PCA
```

# 4. Principal Component Analysis on Continuous Variables

```{r pca_continuous}
pc1 <- prcomp(dcon, scale = TRUE)
class(pc1)
attributes(pc1)

print(pc1)
str(pc1)
```

# 5. Inertia and Variance Percentage

```{r inertia}
# Calculate eigenvalues and inertia
pc1$sdev
inerProj <- pc1$sdev^2 
inerProj
totalIner <- sum(inerProj)
totalIner
pinerEix <- 100 * inerProj / totalIner
pinerEix

# Plot individual explained variance
barplot(pinerEix)

# Plot cumulative explained variance
barplot(100 * cumsum(pc1$sdev[1:dim(dcon)[2]]^2) / dim(dcon)[2])
percInerAccum <- 100 * cumsum(pc1$sdev[1:dim(dcon)[2]]^2) / dim(dcon)[2]
percInerAccum
```

# 6. Selection of Significant Dimensions

```{r select_dimensions}
# Choose first 4 components (~80% total variance)
nd <- 4
print(pc1)
attributes(pc1)
pc1$rotation
```

# 7. Projections in the New Space

```{r projections}
View(pc1$x)
dim(pc1$x)
dim(dcon)
dcon[2000,]
pc1$x[2000,]
# Store projections
Psi <- pc1$x[, 1:nd]
dim(Psi)
# Example access to projection of a single observation
Psi[2000,]

# Labels and indices
iden <- row.names(dcon)
etiq <- names(dcon)
ze <- rep(0, length(etiq))
```

# 8. Plotting Individuals in the First Plane

```{r plot_individuals}
eje1 <- 1
eje2 <- 2

# Plot with labels
plot(Psi[, eje1], Psi[, eje2])
text(Psi[, eje1], Psi[, eje2], labels = iden, cex = 0.5)
axis(side = 1, pos = 0, labels = FALSE, col = "cyan")
axis(side = 3, pos = 0, labels = FALSE, col = "cyan")
axis(side = 2, pos = 0, labels = FALSE, col = "cyan")
axis(side = 4, pos = 0, labels = FALSE, col = "cyan")

# Empty plot with only text
plot(Psi[, eje1], Psi[, eje2], type = "n")
text(Psi[, eje1], Psi[, eje2], labels = iden, cex = 0.5)
axis(side = 1, pos = 0, labels = FALSE, col = "cyan")
axis(side = 3, pos = 0, labels = FALSE, col = "cyan")
axis(side = 2, pos = 0, labels = FALSE, col = "cyan")
axis(side = 4, pos = 0, labels = FALSE, col = "cyan")
```

# 9. Projection of Continuous Variables on the Principal Plane

```{r plot_variables}
Phi <- cor(dcon, Psi)
View(Phi)
X <- Phi[, eje1]
Y <- Phi[, eje2]

plot(Psi[, eje1], Psi[, eje2], type = "n")
axis(side = 1, pos = 0, labels = FALSE)
axis(side = 3, pos = 0, labels = FALSE)
axis(side = 2, pos = 0, labels = FALSE)
axis(side = 4, pos = 0, labels = FALSE)
arrows(ze, ze, X, Y, length = 0.07, col = "blue")
text(X, Y, labels = etiq, col = "darkblue", cex = 0.7)

# Zoom using variable-based limits
plot(Psi[, eje1], Psi[, eje2], type = "n",
     xlim = c(min(X, 0), max(X, 0)),
     ylim = c(-1, 1))
axis(side = 1, pos = 0, labels = FALSE)
axis(side = 3, pos = 0, labels = FALSE)
axis(side = 2, pos = 0, labels = FALSE)
axis(side = 4, pos = 0, labels = FALSE)
arrows(ze, ze, X, Y, length = 0.07, col = "blue")
text(X, Y, labels = etiq, col = "darkblue", cex = 0.7)
```

# 10. Projection of Qualitative Variables (Illustrative Category)

```{r qualitative_illustrative}
# Example with variable in column 16
dd$cap_gain_tocada <- ifelse(dd$cap_gain != 0, 1, 0)
dd$cap_loss_tocada <- ifelse(dd$cap_loss != 0, 1, 0)
names(dd)
varcat <- factor(dd[, 16])
plot(Psi[, 1], Psi[, 2], col = varcat)
axis(side = 1, pos = 0, labels = FALSE, col = "darkgray")
axis(side = 3, pos = 0, labels = FALSE, col = "darkgray")
axis(side = 2, pos = 0, labels = FALSE, col = "darkgray")
axis(side = 4, pos = 0, labels = FALSE, col = "darkgray")
legend("bottomleft", levels(factor(varcat)), pch = 1, col = varcat, cex = 0.6)

k <- 16 # gender (male or female)
varcat <- factor(dd[, k])
fdic1 <- tapply(Psi[, eje1], varcat, mean)
fdic2 <- tapply(Psi[, eje2], varcat, mean)
text(fdic1, fdic2, labels = levels(varcat), col = "yellow", cex = 0.7)
```

# 11. Projection of Centroids for a Qualitative Variable without Individuals

```{r qualitative_centroids}
# Empty plot
plot(Psi[, eje1], Psi[, eje2], type = "n")
axis(side = 1, pos = 0, labels = FALSE, col = "cyan")
axis(side = 3, pos = 0, labels = FALSE, col = "cyan")
axis(side = 2, pos = 0, labels = FALSE, col = "cyan")
axis(side = 4, pos = 0, labels = FALSE, col = "cyan")

# Same variable k = 8
# fdic1, fdic2 assumed computed above
text(fdic1, fdic2, labels = levels(varcat), col = "blue", cex = 0.7)
```

# 12. Joint Projection of All Qualitative Variables

```{r qualitative_joint}
x_center <- mean(Psi[, eje1])
y_center <- mean(Psi[, eje2])

plot(Psi[, eje1], Psi[, eje2], type = "n",
     xlim = x_center + c(-2, 2),
     ylim = y_center + c(-1, 1))
axis(side = 1, pos = 0, labels = FALSE, col = "cyan")
axis(side = 3, pos = 0, labels = FALSE, col = "cyan")
axis(side = 2, pos = 0, labels = FALSE, col = "cyan")
axis(side = 4, pos = 0, labels = FALSE, col = "cyan")

dcat <- c(2:7, 11,12)
colors <- rainbow(length(dcat))
c <- 1
for (k in dcat) {
  seguentColor <- colors[c]
  fdic1 <- tapply(Psi[, eje1], dd[, k], mean)
  fdic2 <- tapply(Psi[, eje2], dd[, k], mean)
  text(fdic1, fdic2, labels = levels(factor(dd[, k])),
       col = seguentColor, cex = 0.6)
  c <- c + 1
}
legend("bottomleft", names(dd)[dcat], pch = 1, col = colors, cex = 0.6)

```


# 13. Custom Color Palette and Final Plot

```{r final_plot}
colors <- c("red", "blue", "darkgreen",
            "orange", "violet", "magenta", "pink")

plot(Psi[, eje1], Psi[, eje2], type = "n", xlim = c(-1, 1), ylim = c(-3, 1))
axis(side = 1, pos = 0, labels = FALSE, col = "cyan")
axis(side = 3, pos = 0, labels = FALSE, col = "cyan")
axis(side = 2, pos = 0, labels = FALSE, col = "cyan")
axis(side = 4, pos = 0, labels = FALSE, col = "cyan")

arrows(ze, ze, X, Y, length = 0.07, col = "lightgray")
text(X, Y, labels = etiq, col = "gray", cex = 0.7)

c <- 1
for (k in dcat) {
  seguentColor <- colors[c]
  fdic1 <- tapply(Psi[, eje1], dd[, k], mean)
  fdic2 <- tapply(Psi[, eje2], dd[, k], mean)
  text(fdic1, fdic2, labels = levels(factor(dd[, k])),
       col = seguentColor, cex = 0.6)
  c <- c + 1
}
legend("bottomleft", names(dd)[dcat], pch = 19, col = colors, cex = 0.6)
```




